---
title: "Awesome R report into bed occupancy and los"
date: "`r Sys.Date()`"
author: S Wellesley-Miller
date-format: "D MMMM YYYY"

format:
  html:
    toc: true
    toc_float: true
    number-sections: false
    title-block-banner: true
    title-block-banner-color: <#005EB8>
    self-contained: true 
---

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri(paste0("NHS_logo.jpg")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:5%; padding:10px;',
               width = "300px",
               heigth = "200px")
```


```{r}
#| echo: false
#| warning: false

library(ggpubr)
library(tidyverse)
library(NHSRplotthedots)
library(gghighlight)
library(geomtextpath)
library(scales)

options(warn = -1)

# generate random data

# ward names
wards <- c("bulbasaur", "squirtle", "charmander", "pikachu")
# number of beds in each ward
no_beds <- c(67, 16, 38, 40)

# number of beds dataframe
beds_data <- data.frame(
  ward = wards,
  no_beds
)

# create a pretty list of names
format_list <- function(x) {
  if (length(x) == 1) {
    return(x)
  }
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

# create random patient data
ward <- sample(c("bulbasaur", "bulbasaur", "squirtle", 
                 "squirtle", "charmander", "pikachu", "pikachu"), 
               3000, replace = TRUE)

# create patiend id number
patient <- seq(1, 3000)

start_date <- as.Date("2025-01-01")
end_date <- as.Date("2025-12-31")

# Generate a sequence of all daily dates in that 12-month window
all_dates <- seq(from = start_date, to = end_date, by = "day")

# Create a data frame, group by month, and sample 35 dates per group
# to ensure there are some each month
random_dates_df <- data.frame(admission_date = all_dates) |>
  mutate(
    Year = year(admission_date),
    Month = month(admission_date, label = TRUE, abbr = FALSE) # Returns full month name
  ) |>
  group_by(Year, Month) |>
  slice_sample(n = 35, replace = TRUE) |>
  arrange(Year, Month, admission_date) |> # Sort chronologically
  ungroup() |>
  slice_sample(n = 3000, replace = TRUE) |> # select 3000 again at random from the start dates
  select(admission_date)

# join all into a single dataframe
ward_data <- bind_cols(
  ward = ward,
  patient = patient,
  random_dates_df
)

# now calulate random end dates
ward_data <- ward_data |>
  rowwise() |>
  mutate(discharge_date = case_when(ward == "bulbasaur" ~ admission_date + days(as.integer(abs(rnorm(1, 20, 4)))),
    ward == "squirtle" ~ admission_date + days(as.integer(abs(rexp(1, 1 / 3)))),
    ward == "charmander" ~ admission_date + days(as.integer(rexp(1, 1 / 5))),
    ward == "pikachu" ~ admission_date + days(as.integer(rexp(1, 1 / 10))),
    .default = admission_date
  )) |>    # charmander ward has some extra stuff happening to it after June
  mutate(discharge_date = if_else(ward == "charmander" & admission_date > as.Date("2025/06/01"),
    discharge_date + as.integer(abs(rnorm(1, 20, 4))),
    discharge_date
  )) |>
  ungroup()

```

```{r} 
#| echo: false
#| warning: false

# start of analysis
ward_data <- ward_data |>
  mutate(los = as.numeric(discharge_date - admission_date))

density_plot <- ward_data |> 
  ggdensity(
  x = "los",
  add = "mean", 
  facet.by = "ward",
  rug = TRUE,
  color = "ward", 
  fill = "ward"
) +
  facet_wrap(~ward,
             scales = 'free_y') +
  theme(legend.position = "none") 

# create spc
ward_data_average <- ward_data |>
  mutate(discharge_month = as.Date(format(discharge_date, "%Y-%m-01")) + 
           months(1) - 1) |>
  summarise(mean_av = mean(los),
            median_av = median(los),
            .by = c(ward,
                    discharge_month)) |>
  arrange(ward,
          discharge_month)

spc_los <- ward_data_average |>
  filter(discharge_month < as.Date('2025/12/31')) |>
  ptd_spc(date_field = discharge_month,
          value_field = median_av,
          facet_field = ward,
          improvement_direction = "decrease",
          target = ptd_target('bulbasaur' = 23,
                    'charmander' = 21,
                    'pikachu' = 15,
                    'squirtle' = 5))

increase <- spc_los |>
  filter(x == max(x),
         point_type == 'special_cause_concern') |>
  pull(f)

spc_los_plot <- spc_los |>
  ptd_create_ggplot(fixed_y_axis_multiple = FALSE,
                    icons_position = "top left",
                    point_size = 2,
                    x_axis_date_format = "%b %y",
                    x_axis_label = 'Discharge month',
                    y_axis_label = 'Median in Days') +
  labs (title = paste0(str_to_title(increase), ' is showing a special cause increase in length of stay'),
        subtitle = paste0('SPC chart of lengths of lengths of stay: ', 
                          format(min(spc_los$x), '%b %y'), 
                          ' - ',
                          format(max(spc_los$x), '%b %y')),
        caption = paste0('Data made up on ',format(Sys.Date(), '%y-%m-%d')))

# calculate occupancy
month_ends <- tibble(month_end = seq(from = start_date, 
                                     to = end_date, 
                                     by = "1 month") |>
                          ceiling_date("month") - days(1))

# join and count patients present at each month-end
monthly_ward_counts <- month_ends |>
  inner_join(ward_data, 
             join_by(between(month_end, 
                             admission_date, 
                             discharge_date))) |> 
  summarise(patient_count = n(), 
            .by = c(ward, 
                    month_end))

# now add on the number of beds and calculate % occupancy

monthly_ward_counts <- monthly_ward_counts |>
  left_join(beds_data) |>
  mutate(perc_occ = patient_count / no_beds)

monthly_ward_counts_plot <- monthly_ward_counts |>
  ggplot() +
  aes(x = month_end,
      y = perc_occ,
      colour = ward) +
  geom_line(size = 1) +
  facet_wrap(~ward) +
  theme_minimal() +
  theme(legend.position = 'none') +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = 'Position at month end',
       y = '%Occupancy',
    title = 'Charmander ward is showing an marked increase in occupancy',
            subtitle = paste0('Line chart of percentage occupancy: ', 
                          format(min(spc_los$x), '%b %y'), 
                          ' - ',
                          format(max(spc_los$x), '%b %y')),
        caption = paste0('Data made up on ',format(Sys.Date(), '%y-%m-%d')))

## charmader focus
latest_full_mth <- max(monthly_ward_counts$month_end)
start <- min(monthly_ward_counts$month_end)
intervention_date <- as.Date("2025-06-30")

y_min <- min(monthly_ward_counts$perc_occ[monthly_ward_counts$ward == "charmander" & monthly_ward_counts$month_end >= intervention_date])

y_max <- max(monthly_ward_counts$perc_occ[monthly_ward_counts$ward == "charmander" & monthly_ward_counts$month_end >= intervention_date])

start <- min(monthly_ward_counts$month_end)

int_dat_less_one <- as.Date("2025-05-30")

y_min_pre <- min(monthly_ward_counts$perc_occ[monthly_ward_counts$ward == "charmander" & monthly_ward_counts$month_end <= int_dat_less_one])
y_max_pre <- max(monthly_ward_counts$perc_occ[monthly_ward_counts$ward == "charmander" & monthly_ward_counts$month_end <= int_dat_less_one])

charmander_focus <- monthly_ward_counts |> 
  ggplot() +
  aes(
  x = month_end,
  y = perc_occ,
  colour = ward
) +
  geom_line(size = 2) +
  scale_x_date(
    date_labels = "%b %y",
    date_breaks = "2 month"
  ) +
  gghighlight(ward == "charmander",
    line_label_type = "sec_axis",
    label_params = list(size = 4)
  ) +
  theme_minimal() +
  annotate("rect",
    xmin = intervention_date, xmax = latest_full_mth,
    ymin = y_min, ymax = y_max,
    fill = "lightblue", alpha = 0.5
  ) +
  annotate("rect",
    xmin = start, xmax = int_dat_less_one,
    ymin = y_min_pre, ymax = y_max_pre,
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_vline(
    xintercept = intervention_date,
    color = "#009639", linetype = "dashed", size = 1
  ) +
  geom_text(
    aes(
      x = intervention_date - months(1),
      y = 0.5, label = "Start of intervention"
    ),
    colour = "#009639", size = 3.5
  ) +
  geom_text(
    aes(
      x = intervention_date + months(3) ,
      y = 0.9, label = "Marked improvement in bed utilisiation"
    ),
    colour = "#005EB8", size = 4, alpha = 0.9
  ) +
  geom_text(
    aes(
      x = intervention_date - months(3),
      y = 0.2, label = "Pre intervention"
    ),
    colour = "#425563", size = 3.5, alpha = 0.7
  ) +
  labs(
    title = "Charmander Ward has improvement in bed utilisation since intervention",
    subtitle = "Chart showing % occupancy Jan 25 - Dec 25 by ward - Charmander Highlighted",
    caption = paste0('Data made up on ',format(Sys.Date(), '%y-%m-%d'))
  ) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

ward_data_joined <- ward_data_average |>
  left_join(monthly_ward_counts,
            by = c('ward' = 'ward', 'discharge_month' = 'month_end'))

ward_data_char <- ward_data_joined |>
  filter(ward == 'charmander',
         !is.na(no_beds)) |>
  mutate(after_intervention = if_else(discharge_month < as.Date('2025-06-01'),
                                      '0',
                                      '1'))

all_wards_corrleation <- ward_data_joined |>
  filter(median_av < 500) |>
  ggplot() +
  aes(x= median_av,
      y= perc_occ) +
  geom_point() + 
  geom_smooth(method = 'lm', se = F) +
  stat_cor(method = 'pearson',
           p.accuracy = 0.001,
           r.accuracy = 0.01) +
  facet_wrap(~ward,
             scales = 'free_x') +
      scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_minimal()+ 
  labs(
    title = "Charmander Ward is showing higher lengths of stay along side higher occupancy since intervention",
    subtitle = "Chart showing mediand LOS and % Occupancy  Jan 25 - Dec 25 by ward",
    caption = paste0("Data made up on ", format(Sys.Date(), "%y-%m-%d")),
    x = "Median length of stay",
    y = "Percent occupancy"
  ) +
  theme(
    legend.position = "none"
  ) 


charmander_corr <- ggplot() +
  geom_point(
    data = ward_data_char,
    aes(
      x = median_av,
      y = perc_occ,
      colour = after_intervention
    )
  ) +
  geom_smooth(
    data = ward_data_char,
    aes(
      x = median_av,
      y = perc_occ
    ),
    method = "lm", se = F
  ) +
  stat_cor(
    method = "pearson",
    p.accuracy = 0.001,
    r.accuracy = 0.01
  ) +
  theme_minimal() +
  labs(
    title = "Charmander Ward is showing higher lengths of stay along side higher occupancy since intervention",
    subtitle = "Chart showing mediand LOS and % Occupancy  Jan 25 - Dec 25 by ward",
    caption = paste0("Data made up on ", format(Sys.Date(), "%y-%m-%d")),
    x = "Median length of stay",
    y = "Percent occupancy"
  ) +
  theme(
    legend.position = "none"
  ) +
  geom_text(
    aes(
      x = 12,
      y = 0.9, label = "Higher utilisation and LOS"
    ),
    colour = "#005EB8", size = 4, alpha = 0.9
  ) 




```



# Introduction

This is a report that explores the distribution of patients lengths of stay and ward occupancy 
across **`r format_list(str_to_title( wards))`** wards.

It contains a bunch of dummy data that is created in the report and specifically Charmander ward has a changed distribution of lengths of stay that occurs in June 2025.

Using the same functions as before and with a few additional elements to the graphs we can combine these elements into an interactive 'Quarto' report.

::: {.callout-note collapse="false" appearance="default" icon="true"}
## Information 

The output of this report is a stand alone HTML file that can be emailed, saved to sharepoint and used like any other file, it can also be opened an utilised on any device such as a pc, tablet or mobile phone.  
:::


# Length of stay

These graphs show distrubtions of LOS over time and the change of median LOS at month end.

::: {.panel-tabset .nav-pills}

### Relative distributions of length of stay by ward


```{r}
#| echo: false
#| warning: false

density_plot

```

This chart allows us to see the expected distribution of length of stay across each ward.  For a ward that does a specific procedure we would perhaps expect to see a bell curve 'normal distribution'.  For a ward that potentially has shorter stays before moving patients on such as an ICU, we may expect to see more of an exponential distribution.  

### SPC of length of stay time series

```{r}
#| echo: false
#| warning: false

spc_los_plot

```

Due to the LOS, not being a normal distribution, median lengths of stay are plotted.

:::

# Occupancy

::: {.panel-tabset .nav-pills}

### Occupancy across all wards

```{r}
#| echo: false
#| warning: false

monthly_ward_counts_plot

```

Exploring all wards we can see that they are mostly in the same range, except for Charmander Ward.  We know that Charmander Ward has had a change in practice as of June 2025 and the focus chart shows more detail on this.  

### Focus on Charmander Ward

```{r}
#| echo: false
#| warning: false

charmander_focus

```

Since the intervention we can clearly see that Charmander Ward has had an increase in occupancy. 

:::

# Correlation of occupancy and lengths of stay

::: {.panel-tabset .nav-pills}

### All wards

```{r}
#| echo: false
#| warning: false

all_wards_corrleation

```

Looking across all wards, only in Charmander ward is there a significant correlation between occupancy and LOS.  

### Focus on Charmander Ward

```{r}
#| echo: false
#| warning: false
charmander_corr

options(warn = 0)

```

We can see that the intervention has resulted in higher occupancy and in addition to a significant increase in length of stay.  

:::

::: {.callout-warning collapse="false" appearance="default" icon="true"}
## Conclusion

This analysis shows that the intervention in Charmander Ward has had an impact on lengths of stay and bed occupancy. 
:::

# Confidentiality 

All data used in this report is complete made up gibberish.  

The code to produce this report is freely available under MIT licence at  

[https://github.com/Simon-W-M/DACOE_demo](https://github.com/Simon-W-M/DACOE_demo)

> Contact  
> [**Simon Wellesley-Miller**](mailto:Simon.Wellesley-Miller@nhs.net)\
> *Senior Analytical Manager*  
>   
> South West Insights and Intelligence
> NHS England  
> Version 1.0 - 15 June 2026


